---
title: "ENVS193DS-Final"
author: "Megan Zhao"
date: 6/9/26
format: pdf
toc: true
editor: source
messages: false
warning: false
---

# Repository Link

[GitHub Repository](https://github.com/meandmymonky/ENVS-193DS_spring-2026_final)

# Set Up

Packages:

```{r}
#| label: reading-packages

# reading in packages
library(tidyverse)
library(here)
library(janitor)
library(readxl)
library(here) 
library(scales) 
library(ggeffects) 
library(MuMIn)
library(forcats)
library(DHARMa)
library(ggeffects)
library(gtsummary)

```

Data:

```{r}
#| label: reading-data

# creating object from 'nests' from data
nests <- read.csv(here('data', 'nest_data_final.csv'))

# creating object 'mz_data' from data
mz_data <- read_xlsx(here('data', '193DS-personal-data-updated.xlsx'))

```

```{r}
#| label: cleaning-personal-data
#| echo: false

# create new object called 'mz_clean' from 'mz_data'
mz_clean <- mz_data |> 
  # clean names
  clean_names() |>
  # create new data column 'completed_v_intended'
  mutate(
    # use data from select columns to calculate new column
    completed_v_intended = (
      number_of_completed_tasks/number_of_intended_tasks
    ) * 100) |> 
  # use data from select columns to calculate new column
  mutate(total_activity_time = (
    time_spent_on_activites_hrs + time_spent_in_class_hrs
    )) |> 
  select(
    date,
    completed_v_intended,
    total_activity_time
)

```


# Problem 1

## a.

In part 1 of the results section, the co-worker used a linear regression. In part 2 of the results section, the co-worker used a Chi-Squared test. 

## b.

Statement 1: Based collected American Avocet micro-habitat use and distance to water edge, American Avocet micro-habitat use is significantly influenced by proximity to water edge. 

Statement 2: Based on collected shorebird species presence across the habitats managed pond, non-tidal marsh, salt pond, tidal marsh, and tidal mudflap, shorebird species significantly differ by locale. 

## c.

One additional variable that could influence the probability of American Avocet microhabitat use is the water depth surrounding habitats, a continuous variable measured in meters, collected via bathymetry or hands-on-measurement depending on depth. This variable might influence American Avocet habitat use as water depth is a significant driver of bird use in managed wetlands -- water depth may influence American Avocet use depending on feeding patterns (deeper areas are inaccessible to wading birds), present food type (different depths may lead to different flora and fauna presence), and predator presence (predators may reside in varying depths of water/predators may not be able to access certain water depths), playing a role in determining how American Avocets use these locales (Schacter et. al., 11). 

# Problem 2

## a.

```{r}
#| label: cleaning-data

# create new object called 'nests_clean' from 'nests'
nests_clean <- nests |> 
  # clean names
  clean_names() |> 
  # filter ant species
  filter(
    ant_species %in% c("RRB", "AB", "BBR")) |> 
  # full names
  mutate(
    species_name_full = recode_values(ant_species,
                                      "RRB" ~ "Crematogaster mimosae",
                                      "AB" ~ "Crematogaster sjostedti", 
                                      "BBR" ~ "Crematogaster nigriceps")) |> 
  # reordering species
  mutate(
    # order of species
    species_name_full = as_factor(species_name_full),
    species_name_full = fct_relevel(species_name_full,
                                    "Crematogaster mimosae",
                                    "Crematogaster sjostedti",
                                    "Crematogaster nigriceps")) |>
  # arrange in order
  arrange(species_name_full) |> 
  # columns of interest
  select(
    height_cm, 
    case_control, 
    species_name_full)

# display structure
str(nests_clean)
# display 10 random rows
nests_clean |> slice_sample(n = 10)

```

## b. 

In this data set, 1s and 0s indicate if a tree has a nest (1) or is "available" (0); in other words, they indicate if a tree contains a birds nest or not. 

## c.

Tree Height: Tree height is a continuous variable, measured in centimeters. According to paragraph three in the introduction subsection, as well as paragraph two in the results subsection of "Symbiotic acacia ants drive nesting behavior by birds in an African savanna", as tree height increases, probability of nest occurrence probability will increase as well, as higher nests decrease ground predator nest predation rates, increasing overall bird lifetime fitness. 

Acacia Ant Species: Acacia ant species is a categorical variable (C.nigriceps, C.mimosae, and C.sjostedti). According to paragraph three in the introduction subsection in "Symbiotic acacia ants drive nesting behavior by birds in an African savanna", the presence of more aggressive defending symbionts (C.nigriceps and C.mimosae) will result in an increased probability of nest occurrence probability, aggressive defender presence may protect nests, reducing nest predation rates, and increase overall bird lifetime fitness.

## d.

|  Model number | Ant Species | Tree Height | Predictor List                   |  
|:-------------:|:-----------:|:-----------:|:--------------------------------:|  
| 0 (null)      |             |             |  no predictors                   |
| 1 (full model)|      x      |      x      |  all predictors                  | 
| 2 (species)   |      x      |             |  species                         |
| 3 (height)    |             |      x      |  tree height                     |
| 4 (interact)  |      x      |      x      | all predictors -- reliance       |

## e.

```{r}
#| label: model-running
#| output: false

# model 0: null model
model_0 <- glm(
  # formula
  case_control ~ 1,
  # data frame
  data = nests_clean,
  family = "binomial") 

# model 1: no interaction
model_1 <- glm(
  # formula
  case_control ~ height_cm + species_name_full,
  # data frame
  data = nests_clean,
  family = "binomial") 

# model 2: ant species
model_2 <- glm(
  case_control ~ species_name_full,
  # data frame
  data = nests_clean,
  family = "binomial") 

# model 3: tree height
model_3 <- glm(
  # formula
  case_control ~ height_cm,
  # data frame
  data = nests_clean,
  family = "binomial") 

# model 4: interaction
model_4 <- glm(
  # formula
  case_control ~ height_cm * species_name_full,
  # data frame
  data = nests_clean,
  family = "binomial") 

```


## f.

```{r}
#| label: AIC

# running AICc
AICc(
  model_0,
  model_1,
  model_2,
  model_3,
  model_4
) |> 
  # display ordered results
  arrange(AICc)

```

The best model as determined by Akaike’s Information Criterion (AIC) was the model using interacting predictor variables ant species and tree height for the response variable of bird nest presence (model 4).

## g.

```{r}
#| label: dharma-diagnostics

# running dharma diagnostic
diagnostic_checking <- simulateResiduals(fittedModel = model_4)

# display dharma diagnostic
plot(diagnostic_checking)
```
```{r}
#| label: summarizing
#| echo: false
#| output: false

# display summary
summary(model_4)
```


## h.

```{r}
#| label: all-predictions

# generating all predictions
predictions <- ggpredict(
  model_4,
  terms = c("height_cm [all]", 
            "species_name_full")) |> 
  # renaming columns
  rename(height_cm = x,
         species_name_full = group)

```

```{r}
#| label: data-visualization

# base layer: ggplot
# use nests_clean data
ggplot(data = nests_clean,
       # x-axis
       aes(x = height_cm,
           # y-axis
           y = case_control)) +
  # first layer: points
  # color by species
  geom_point(aes(color = species_name_full),
             # change point shape
             shape = 21) +
  # second layer: line
  # use predictions data
  geom_line(data = predictions,
            # x-axis
            aes(x = height_cm,
                # y-axis
                y = predicted,
                # color by species
                color = species_name_full),
            # set linewidth
            linewidth = 2) +
  # third layer: ribbon
  # use predictions data
  geom_ribbon(data = predictions,
              # x-axis
              aes(x = height_cm,
                  # y-axis
                  y = predicted,
                  # y-min
                  ymin = conf.low,
                  # y-max
                  ymax = conf.high,
                  # fill by species
                  fill = species_name_full),
              # change opacity
              alpha = 0.1) +
  # panels by species
  facet_wrap(vars(species_name_full)) +
  labs(
    # label x-axis
    x = "Tree Height (cm)",
    # label y-axis
    y = "Probability of nest occurence (%)") + 
  # color by species
  scale_color_manual(
    values = c("Crematogaster mimosae" = "violetred2",
               "Crematogaster sjostedti" = "steelblue1",
               "Crematogaster nigriceps" = "mediumorchid4")) +
  # use classic theme
  theme_classic() +
  # remove legend
  theme(legend.position = "none")

```


## i.

*Figure 1. Ant species tree presence and tree height on nest presence probability.* Points represent observations of individual nest occurrences (0||1) with their respective tree heights (cm). Solid lines represent model predictions for nest presence, with shaded areas indicating the 95% confidence interval range. Colors represent ant species (pink: C. mimosae, blue: C. sjostedti, purple: C. nigriceps). Data originally from Lujan, E., Nielsen, R., Short, Z., Wicks, S., Nderitu Watetu, W., Malingati Khasoha, L., Palmer, T. M., Goheen, J. R., & Alston, J. M. (2023). Data, code, and metadata for: **Symbiotic acacia ants drive nesting behavior by birds in an African savanna** [Data set]. Zenodo. https://doi.org/10.5281/zenodo.8373322

## j.

```{r}
#| label: predictions

# generating prediction
ggpredict(model_4,
          # setting terms
          terms = c("height_cm [600]",
                    "species_name_full"))

```

## k.

Ant species presence and tree height interact to predict probability of bird presence. As ant aggressiveness and tree height increase, the probability of nest presence increases. At 600 cm, nest presence probability is 0.74 (95% CI:[0.48, 0.90]) for C. mimosae, 0.20 (95% CI:[0.00, 0.99]) for C. sjostedti, and 1.00 (95% CI:[1.00, 1.00]) for C. nigriceps; bird nest occurrence occurs at higher rates in trees with more aggressive acacia ant species, indicating that more aggressive symbionts may confer additional protections from predators beyond residing trees to bird nests, reducing nest predation rates and increasing overall bird lifetime fitness.

# Problem 3

## a. 

My visualizations used different metrics and different model types across Homework 2/3 and my affective visualization; in Homework 2 and 3 I used box plots to visualize my data, while I used a trend line to visualize my data in my affective visualization. Across my visualizations, I did not see many similarities -- my affective visualization reflected my predicted outcomes with different variables, while my visualizations in homework 2/3 did not show the trends that I depicted in my affective visualization; similarly, I did not see many patterns across my affective and effective visualizations, likely due to different metrics of creation and data alignment. I did not receive any feedback on my Homework 2 and 3 visualizations, however, in workshop, I got a lot of visual suggestions -- for instance, implementing a clearer trend line in my drawing -- which was reflected in my final affective visualization (I added a red trend line, and used colors to create clearer visual distinctions between my data). 

## b. 

Using the response variable of productivity % and the predictor variable of activity time (as expressed in my affective visualization as an aggregation of outside activity and class time), one analysis that we talked about this quarter that I could apply to my research question is a correlation analysis, specifically a Spearmans correlation. Both of my variables are independent numerical continuous variables, with one variable (activity time) predicted to have an influence on the other variable (productivity %), suspected monotonic relationship. Additionally, there are not that many observations in data. Subsequently, I believe that a Spearmans correlation analysis is appropriate in trying to answer the question: how do my activities impact my productivity?

## c.

```{r}
#| label: assumption-checking

# base layer: ggplot
ggplot(
  # starting data frame
  data = mz_clean,
  # x-axis
  mapping = aes(x = total_activity_time,
                # y-axis
                y = completed_v_intended)
) +
  # first layer: points
  geom_point(
    # coloring points
    color = "chartreuse4",
    # changing opacity
    alpha = 0.8
  ) +
  # relabelling x-axis, y-axis, and title
  labs(
    # label x-axis
    x = "Time Spent on Activites (hours)",
    # label y-axis
    y = "Tasks Completed (%)",
    # add chart title
    title = 'Productivity and Activity Time Correlation') +
  # use classic theme
  theme_classic() +
  # remove legend
  theme(legend.position = "none")

```

```{r}
#| label: analysis-running

cor.test(
  # y-variable
  mz_clean$completed_v_intended,
  # x-variable
  mz_clean$total_activity_time,
  # use spearmans test
  method = 'spearman',
  # use normal p-approximation
  exact = FALSE)

```

## d.

```{r}
#| label: spearman-visualization

# base layer: ggplot
ggplot(
  # starting data frame
  data = mz_clean,
  # x-axis
  mapping = aes(x = total_activity_time,
                # y-axis
                y = completed_v_intended)
) +
  # first layer: points
  geom_point(
    # coloring points
    color = "chartreuse4",
    # changing opacity
    alpha = 0.8
  ) +
  # relabelling x-axis, y-axis, and title
  labs(
    # label x-axis
    x = "Time Spent on Activities (hours)",
    # label y-axis
    y = "Tasks Completed (%)",
    # add chart title
    title = 'Activity Time v. Productivity') +
  # use classic theme
  theme_classic() +
  # remove legend
  theme(legend.position = "none")

```
## e.

*Figure 2. Activity time decreases productivity.* Small round solid green points represent individual measurements of productivity (% of tasks completed) in relation to time spent that day on activities (hours). Data from `193DS-personal-data-updated` (Zhao M, 2026).

## f.

To evaluate the strength of the relationship between activity time and productivity I used a Spearmans rank correlation test. I chose to use Spearmans test over Pearsons as I did not check for normality nor a linear relationship. After running the Spearmans test, I found a moderate negative relationship between ocean temperature and giant kelp frond elongation rate (Spearman $\rho$ = -0.58, S = 5787.6, p = 0.001, $\alpha$ = 0.05). As time spent on activites increases, productivity decreases.

## g.

Completed in section